---
title: "Unsupervised Breast Cancer Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Unsupervised Breast Cancer Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 7,
  dev = "ragg_png"
)
```

In this analysis we will cluster breast cancer samples from TCGA based 
on gene expression and explore their sub-type.

## Loading required packages:

```{r setup}
library(XIFF)
library(TIFF)
library(dplyr)
library(ggplot2)
```

## Database connection, tissue and gene annotation

First we load the settings for the database connection and
set the respective options to allow a smooth db connection.

Then we filter the tissue annotation to breast cancer samples
and load the gene annotation.

```{r}
setDbOptions(getSettings())
set.seed(123)

tissue_anno <- getTissueAnnotation() %>%
  filter(tissuepanel == "TCGA tumors" & tumortype == "breast invasive carcinoma")

gene_anno <- getGeneAnno("human")
```

## Dimension reduction based on a hallmark gene set

Among the available MSigDB Hallmark gene sets we select the genes, which
are involved in the early response to estrogen. 
The variable ensg_gene_set will be a vector of ensembl gene names (ENSG).

Then we retrieve the gene expression (TPMs and COUNTs) for the respective genes
and breast cancer samples, apply two dimension reduction methods (T-SNE by default
and UMAP) and plot the results to see, if they are clustering by histology sub type.

We can see a good separation of the Basal/Triple-Negative samples. The HER2/ERBB2 high 
samples also cluster in one area, but there is hardly any separation of the Luminal 
A and B.

```{r}
my_gene_set <- grep("ESTROGEN_RESPONSE_EARLY", getAvailableHallmarkGeneSets(), value = TRUE)
ensg_gene_set <- getGSEAdata("human", "hallmark", gene_set = my_gene_set)

breast_dimred <- getExpressionDimRedData(tissue_anno$tissuename, ensg_gene_set, p = TRUE)

tissueDimRed <- getExpressionDimRed(data = breast_dimred, anno = tissue_anno, method = "tsne")
generateExpressionDimRedPlot(tissueDimRed, colorCol = "histology_subtype")
```

## Dimension reduction based on selected genes

Selecting a number of breast cancer relevant genes gives us more flexibility to focus 
on the most important genes, rather than a full gene set with ~200 genes.
Beside the hormone receptors for estrogen (ESR1) and progesterone (PGR), we selected 
elastin (ELN), Aurora kinase A (AURKA) and odd-skipped related transcription 
factor 1 (OSR1).

This time the separation between Luminal A and B is much clearer.

```{r}
breast_genes <- gene_anno %>%
  filter(symbol %in% c("ESR1", "PGR", "ERBB2", "AURKA", "ELN", "OSR1") & grepl("ENSG", ensg))

breast_custom_dimred <- getExpressionDimRedData(tissue_anno$tissuename, breast_genes$ensg, p = TRUE)

tissueDimRed2 <- getExpressionDimRed(data = breast_custom_dimred, anno = tissue_anno, 
                                     method = "tsne", clusterMethod = "k-means k = 4")
generateExpressionDimRedPlot(tissueDimRed2, colorCol = "histology_subtype")
```

## Gene expression cluster overlapps with histology type

Subtypes BRCA_Her2 (ERBB2 high samples) and BRCA_Basal (triple negative samples) 
overlap with gene expression clusters.

```{r}
generateExpressionDimRedPlot(tissueDimRed2, colorCol = "cluster")
ggplot(tissueDimRed2, aes(x = histology_subtype, fill = cluster)) + geom_bar()
```

With this vignette we could show a good separation of Breast subtypes 
based on gene expression.
